Solving the maximum independant set problem with einsum networks

### 2021-06-01 paper/paper.tex
independence polynomial
update computing configurations

%\documentclass[a4paper,superscriptaddress,11pt]{quantumarticle}
%\documentclass[aps,twocolumn,longbibliography,english,superscriptaddress]{revtex4-1}
\documentclass{article}
\usepackage{iclr2021_conference}
%\documentclass[a4paper,superscriptaddress,11pt]{article}
\pdfoutput=1
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithm}
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color,amsthm}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

@@

\begin{abstract}
	Solving the maximum independent set size problem: the maximum independent set size, the independence polynomial and the optimal configuration.
\end{abstract}

\section{Computing independence polynomial}
The independence polynomial~\cite{Ferrin2014, Harvey2017} of a graph $G$ is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of order $k$ in $G$, and $\alpha(G)$ is the maximum independent set size of graph $G$.

If we map a graph into an einsum network, as shown in \Fig{fig:einsummapping}, by putting a vertex tensor of size $2$ at each vertex
\begin{equation}
    W(x) = \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right),
\end{equation}
were $x$ is a variable, and a matrix of size $2 \times 2$ at each edge
\begin{equation}
    B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right).
\end{equation}

The contractionof this einsum network is exactly the independence polynomial %, we denote this contraction as \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
\begin{equation}
    I(G, x) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i, s_j}.
\end{equation}
Here, the einsum enumerates all possible configurations and accumulates the product of tensor elements to the output. The product over edge tensors will provide a factor $0$ if the configuration is not an independent set, otherwise $1$.
While the product over edges provide a factor $x^k$, where $k$ depends on how many times a vertex takes value $1$.
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
        \caption{}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
    \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
        \caption{}
    \end{subfigure}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}
The benefit of mapping the problem to einsum is, some existing algorithms can be used to decrease the complexity of the problem.
Especially those techiniques developed in recent quantum circuit simulations for finding the optimal exactly contraction order of a tensor network ~\cite{Gray2021,Pan2021}.
With these algorithms, one can easily find a good contraction order that reduces the complexity to a quantity related to the minimum treewidth of the graph.~\cite{Markov2008}
Such mapping makes it possible to utilize fater linear algebra functions.

@@

\section{Computing maximum independent set size and its degeneracy}
Obtaining the maximum independent set size and its degeneracy is much easier. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x + a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= -\infty\\
    \mymathbb{1} &= 0
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$. This is exactly the tropical tensor network~\cite{Liu2021} used in solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$, we can only consider the powers, which corresponds to the tropical algebra~\cite{Maclagan2015}.

>>>>

%\documentclass[aps,twocolumn,longbibliography,english,superscriptaddress]{revtex4-1}
\documentclass{article}
\usepackage{iclr2021_conference}
%\documentclass[a4paper,superscriptaddress,11pt]{article}
\pdfoutput=1
\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithm}
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color,amsthm}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

@@

\begin{abstract}
	Solving the maximum independent set size problem by mapping the problem to an einsum network. 
    We obtain the maximum independent set size,
    the independence polynomial and the optimal configuration by using different tensor elements to a commutative semiring.
\end{abstract}

\section{Computing independence polynomial}
The independence polynomial~\cite{Ferrin2014, Harvey2017} of a graph $G$ is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of order $k$ in $G$, and $\alpha(G)$ is the maximum independent set size of graph $G$.
If we map a graph into an einsum network, as shown in \Fig{fig:einsummapping}, by putting a vertex tensor of size $2$ at each vertex
\begin{equation}
    W(x) = \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right),
\end{equation}
and a matrix of size $2 \times 2$ at each edge
\begin{equation}
    B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right).
\end{equation}

Here $0$ and $1$ are the additive identity and multiplicative identity of a commutative semiring respectively.
$x$ is a variable.
The contraction of this einsum network is the independence polynomial %, we denote this contraction as \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
\begin{equation}
    I(G, x) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i, s_j}.
\end{equation}
Here, the einsum enumerates all possible configurations and accumulates the product of tensor elements to the output. The product over edge tensors will provide a factor $0$ if the configuration is not an independent set, otherwise $1$.
While the product over edges provide a factor $x^k$, where $k$ depends on how many times a vertex takes value $1$.
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
        \caption{}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
    \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
        \caption{}
    \end{subfigure}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}
The benefit of mapping the problem to einsum is, some existing algorithms can be used to decrease the complexity of the problem.
Especially those techiniques developed in recent quantum circuit simulations for finding the optimal exactly contraction order of a tensor network ~\cite{Gray2021,Pan2021}.
With these algorithms, one can easily find a good contraction order that reduces the complexity to a quantity related to the minimum treewidth of the graph.~\cite{Markov2008}
Such mapping makes it possible to utilize fater linear algebra functions.

@@

\section{Computing maximum independent set size and its degeneracy}
Obtaining the maximum independent set size and its degeneracy is much easier. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= -\infty\\
    \mymathbb{1} &= 0
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$. This is exactly the tropical tensor network~\cite{Liu2021} used in solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$, we can only consider the powers, which corresponds to the tropical algebra~\cite{Maclagan2015}.

One may want to obtain all ground state configurations,
it can be achieved replacing the factors $a_x$ with a set of bit strings $s_x$ initialized to a vertex dependent value $s_i = \{\delta_{ij=1,\ldots,n}$\}.
The bit strings has a redefined algebra
\begin{align}
\begin{split}
    s_x \oplus s_y &= s_x \cup s_y\\
    s_x \odot s_y &= \{\sigma \lor \tau | \sigma \in s_x, \tau \in s_y\}
\end{split}
\end{align}
The $\vee$ is the bit-wise \texttt{or}.
